Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-01-13 3.phase_transitions/phase_transitions.tex
minor edits
added

1) ?.Example.tex files for each chapter conforming to the current style.

2) Began the "index".  If you want to include a term do \index{term}{term}

3) Fixed the level of the table-of-contents.

We study the fundamental limits of the signal detection and support recovery problems in the location models with independent Gaussian errors in this chapter.
Specifically, we derive the conditions under which the detection and support recovery problems succeed and fail in the sense of \eqref{eq:support-recovery-success} and \eqref{eq:support-recovery-failure}, in the additive error model 
\begin{equation} \label{eq:model-additive-Chapter3}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p,
\end{equation}
where the errors $\epsilon(i)$'s are \ac{iid} standard Gaussians random variables.
We restrict our analysis to models with independent and identically distributed Gaussian errors for the moment.
Both the distributional assumption and the independence assumption will be relaxed in the following chapters.

As laid out in Section \ref{sec:asymptotics}, we work under the asymptotic regime where the problem dimension $p$ diverges to infinity.
Let the signal vectors $\mu = \mu_p$ have 
\begin{equation} \label{eq:signal-sparsity-additive}
    |S_p| = \left\lfloor p^{1-\beta} \right\rfloor, \quad \beta\in(0,1]
\end{equation}
non-zero entries, where $\beta$ parametrizes the problem sparsity.
The closer $\beta$ is to 1, the sparser the support $S$; conversely, when $\beta$ is close to 0, the support is dense with many non-null signals.
We consider one-sided alternatives \eqref{eq:global-test-one-sided}, and parametrize the range of the non-zero (and perhaps unequal) signals with
\begin{equation} \label{eq:signal-size-additive}
    \underline{\Delta} = \sqrt{2\underline{r}\log{p}}
    \le \mu(i) \le
    \overline{\Delta} = \sqrt{2\overline{r}\log{p}}, \quad \text{for all}\;\;i\in S_p,
\end{equation}
for some constants $0<\underline{r}\le\overline{r}\le+\infty$.

@@

\section{Proofs}
\label{sec:proofs}
\input{3.phase_transitions/3.proofs-phase-transitions.tex}

>>>>

\stilian{The purpose of this chapter is to provide a unified review of the fundamental statistical limits in the 
sparse signal detection and support recovery problems.  Our goal is to convey the main ideas and thus we shall focus on 
the simple but important setting of independent Gaussian errors.}{\xout{We study the fundamental limits of the signal detection and support recovery problems in the location models with independent Gaussian errors in this chapter.}}
Specifically, we derive the conditions under which the detection and support recovery problems succeed and fail in the sense of \eqref{eq:support-recovery-success} and \eqref{eq:support-recovery-failure}, in the additive error model 
\begin{equation} \label{eq:model-additive-Chapter3}
    x(i) = \mu(i) + \epsilon(i), \quad i=1,\ldots,p,
\end{equation}
where the errors $\epsilon(i)$'s are \ac{iid} standard Gaussians random variables.
Once again, we restrict our analysis to models with independent and identically distributed Gaussian errors for the 
moment. Both the distributional assumption and the independence assumption will be relaxed substantially in the following chapters.

As laid out in Section \ref{sec:asymptotics}, we work under the asymptotic regime where the problem dimension 
$p$ diverges to infinity.  The set of non-zero entires of the signal vector $\mu = \mu_p$ will be referred
to as its {\em support} and denoted by 
$$
 S_p:=\{i\, :\, \mu(i)\not = 0\}.
$$
We shall assume that the size of the support is
\begin{equation} \label{eq:signal-sparsity-additive}
    |S_p| = \left\lfloor p^{1-\beta} \right\rfloor, \quad \beta\in(0,1],
\end{equation}
where $\beta$ parametrizes the problem sparsity.
The closer $\beta$ is to 1, the sparser the support $S_p$.  
Conversely, when $\beta$ is close to 0, the support is dense with many non-null signals.
We consider one-sided alternatives \eqref{eq:global-test-one-sided}, and parametrize the range of the non-zero (and perhaps unequal) signals with
\begin{equation} \label{eq:signal-size-additive}
    \underline{\Delta} = \sqrt{2\underline{r}\log{p}}
    \le \mu(i) \le
    \overline{\Delta} = \sqrt{2\overline{r}\log{p}}, \quad \text{for all}\;\;i\in S_p,
\end{equation}
for some constants $0<\underline{r}\le\overline{r}\le+\infty$.

@@

%\section{Proofs}
%\label{sec:proofs}
%\input{3.phase_transitions/3.proofs-phase-transitions.tex}

### 2021-01-13 6.GWAS/GWAS.tex
Addressed the minor typos of R2
I defined the command "\stilian" in the preamble and used it to make changes in response to the minor typos.

Please see reviews/GS_Annotated.pdf.

All phase transition boundaries coincide with those in the additive error models obtained in Chapter \ref{chap:phase-transitions} under suitable parametrizations.
% and Figure \ref{fig:phase-chi-squared} continues to apply.
indicating vanishing differences between the difficulties of the one-sided and two-sided alternatives in the Gaussian additive error model \eqref{eq:model-additive}.

@@

\section{Exercises}
\label{sec:Exercises-GWAS}
\input{6.GWAS/6.Exercises.tex}

>>>>

All phase transition boundaries coincide with those in the additive error models obtained in Chapter \ref{chap:phase-transitions} under suitable parametrizations.
% and Figure \ref{fig:phase-chi-squared} continues to apply.
This indicates vanishing differences between the difficulties of the one-sided and two-sided alternatives in the Gaussian additive error model \eqref{eq:model-additive}.

@@

### 2021-01-13 preamble.tex

This text presents a collection of new results and recent developments on the phase transition phenomena in sparse signal problems.
The main theme is the study of the fundamental limits in high-dimensional testing and inference. Since the seminal works
of \cite{ingster1998minimax} and \cite{donoho2004higher}, the subject has received a lot of attention in the literature with important contributions
from \cite{ji2012ups,genovese2012comparison,jin2014optimality,arias2017distribution,butucea2018variable}.  These works (among others) 
 have discovered the presence of fundamental limits in the so-called  {\em needle in haystack} problems, where a sparse signal is observed in 
 high-dimensional additive noise.  In this setting, two archetypal  problems arise -- the {\em signal detection} and {\em signal support recovery}.  
 Signal detection is a hypothesis testing problem, which amounts
to testing for the presence of non-zero signal in some (any) of the dimensions.  Support recovery, on the other hand, is an inference problem, which aims to estimate the signal support -- the locations of the non-zero signal components.  The fundamental limits of these problems are studied in the so-called high-dimensional asymptotics regime where the sample size $n$ is fixed and the dimension $p$ of the underlying signal 
grows to infinity.

From a probabilistic perspective, the above phase transition phenomena are asymptotic zero-one type laws, as $p\to\infty$.  
Namely, consider a sparse signal with  {\em support size} on the order of $p^{1-\beta}$ for some parameter $\beta\in (0,1)$.  
Parameterize the non-zero {\em signal amplitude} by $\Delta(p^r)$, 
for some $r>0$ and a suitable monotone non-decreasing function $\Delta(\cdot)$.  Then, for a broad range of error distributions and
statistical problems, one encounters a sharp transition between the regimes where the problem is solvable and unsolvable depending on 
the signal magnitude $r$ and signal sparsity $\beta$.  More precisely, there exists a boundary function $f(\beta)$ such that if 
the signal magnitudes are {\em above} the boundary, $r> f(\beta)$, then the problem can be solved with vanishing loss as $p\to\infty$, with a suitable statistical procedure.  On the other hand, if the signal is below that same boundary, i.e., $r<f(\beta)$, all statistical procedures fail to  provide a solution with a vanishing loss, as $p\to\infty$. Of course, depending on whether one considers the  detection (testing) or 
support recovery (inference) problems, different loss functions quantify success and failure. The choice of the loss functions is often guided by the 
applications, resulting in a rich picture of phase-transitions (see e.g.\ Figure \ref{fig:phase-Gaussian-errors}).

@@

{\bf Acknowledgements.} The authors gratefully acknowledge the support of their families and all colleagues from the Statistics Department 
at the University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, 
Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu, \stilian{R2, Typo 1}{\fbox{add more}}
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.

>>>>

This text presents a collection of new results and recent developments on the phase transition phenomena in sparse signal problems.
The main theme is the study of the fundamental limits in high-dimensional testing and inference. Since the seminal works
of \cite{ingster1998minimax} and \cite{donoho2004higher}, the subject has received a lot of attention in the literature with important contributions
from \cite{ji2012ups,genovese2012comparison,jin2014optimality,arias2017distribution,butucea2018variable}.  These works, among many others, have discovered some fundamental limits in the so-called  {\em needle in haystack} problems, where a sparse signal is observed with high-dimensional additive noise.  In this setting, two archetypal  problems arise -- the {\em signal detection} and {\em signal support recovery}.  
The signal detection refers to a global hypothesis testing problem that amounts
to determining the presence of non-zero signal in any of its dimensions. 
The support recovery, on the other hand, can be seen either as a multiple testing problem where the presence of non-zero signal is tested for each signal location of interest, or alternatively,  as an inference problem that aims to estimate the signal support, i.e., the locations of the non-zero signal components.  The fundamental limits of these problems are studied in the so-called high-dimensional asymptotic regime where the dimension $p$ of the underlying signal grows to infinity, and the sample size $n$ is either bounded or grows slowly relative to $p$.

From a probabilistic perspective, these aforementioned fundamental limits are stated as asymptotic zero-one type laws, as dimensionality diverges.  
Namely, consider a sparse signal with  {\em support size} on the order of $p^{1-\beta}$ for some parameter $\beta\in (0,1)$.  
Parameterize the non-zero {\em signal amplitude} by $\Delta(p^r)$, 
for some $r>0$ and a suitable monotone non-decreasing function $\Delta(\cdot)$.  Then, for a broad range of error distributions and
statistical problems, one encounters a sharp transition between the regimes where the problem is solvable and unsolvable depending on 
the signal magnitude $r$ and signal sparsity $\beta$.  More precisely, there exists a boundary function $f(\beta)$ such that if 
the signal magnitudes are {\em above} the boundary, $r> f(\beta)$, then the problem can be solved with vanishing loss as $p\to\infty$, with a suitable statistical procedure.  On the other hand, if the signal is below that same boundary, i.e., $r<f(\beta)$, all statistical procedures fail to  provide a solution with a vanishing loss, as $p\to\infty$. Of course, depending on whether one considers the  detection (testing) or 
support recovery (inference) problems, different loss functions quantify success and failure. The choice of the loss functions is often guided by the 
applications, resulting in a rich picture of phase-transitions (see e.g.\ Figure \ref{fig:phase-Gaussian-errors}).

@@

{\bf Acknowledgements.} The authors gratefully acknowledge the support of their families and all colleagues from the Statistics Department 
at the University of Michigan, Ann Arbor.   Special thanks (in alphabetical order) go to Xuming He, Tailen Hsing, Michalis Kallitsis, Liza Levina, Yuanzhi Li, Rodderick Little, Ya'acov Ritov, Kerby Shedden, Jinqi Shen, Jonathan Terhorst, Gongjun Xu.
The authors were partially supported by the NSF program {\em Algorithms for Threat Detection}.
